TODO: should mention the name of the language
The original idea for the project was to create a EcmaScript like language for mobile devices. 
Unfortunately during the project, it turned out that such an implementation already exists,
though it is proprietary and only little information is available about it \cite{mojax}.

So there needed to be a change in the project.
The new approach to the project is to focus on another niche within scripting languages on mobile devices. Namely to create a scripting language which runs on even the very low end mobile devices.
The motivation for this is that even scripting languages optimised for mobile devices, such as hecl\cite{hecl}, kahlua\cite{kahlua}, fscriptme\cite{fscriptme}, simkin\cite{simkin}, have very large code footprints compared to the amount of space available on the lowest end mobile devices: ca. 50-200KB code footprint for the scripting language vs. 64K available on low end device for the entire application, including embedded scripting language.
So the new target is to create a scripting language designed for small footprint, while staying highly expressive.

The project creates a scripting language, where the implementation can be, and is done significantly smaller than any other scripting language currently available for a mobile platform.
This comes at the price of having lisp-like syntax and dynamic scoping, but otherwise it is similar to other scripting languages / extension languages. It is also very easy to embed and use as an extension language for other applications.
