This chapter surveys existing languages.

\section{JVM scripting languages resource usage}
This section looks at some of the scripting languages that exists for the JVM,
mainly with focus on the mobile platform and low resource usage.

To compare the resource usage, I have added an estimate of the jar file size. 
The estimate is done by making a zip archive of the non-obscurified class files,
trying to exclude class files that are part of extra libraries.
The motivation for this approach is that some of the languages have large GUI libraries,
which would count against them, if we just looked at the jar file. 
In addition, compilation of some of the languages is obscurified/optimised by default,
which would give those an advantage. So this should give a more fair comparison.
As a rule of thumb with classfiles, the compressed files are approximately half the size of original.
 
\begin{figure}
\begin{center}
\begin{tabular}{|c|r|}
\hline 
Jscheme & 30K \\
\hline 
Kahlua & 40K \\
\hline 
Hecl & 50K \\
\hline 
Simkin & 80K \\
\hline 
CellularBasic & 90K \\
\hline 
Rhino & 400K \\
\hline 
\end{tabular}
\end{center}
\caption{Approximation of jar size of scripting languages for the jvm}
\end{figure}

Looking for scripting languages for Java Mobile Edition, there seem to be Simkin, CellularBasic, Kahlua, FscriptME and Hecl. There is also Mojax, but that is a proprietary system, and thus only little information is availabe about it.
These are compared with JScheme and rhino for perspective.


\subsection{Hecl}
Hecl seems to be the major scripting language for mobile devices.
It has different editions targeting both CLDC1.0, CLDC1.1 and Android.
The design is modular, where libraries can be enabled/disabled.
It is a shrunk down dialect of Tcl. Arithmetic operators are also functions, which mean that calculated expressions must be written Lisp-like: e.g. \verb|[+ 1 2 [* 3 4]]|

\subsection{Lua/kahlua}
\subsection{FscriptME}
Fscript\cite{fscriptme}, and FscriptME as an implementation for Java Micro Edition.
This language
This language is still in beta, 

\subsection{Simkin}
Simkin is an XML scripting language

\subsection{CellularBasic}
This is an implementation of a basic dialect for mobile devices.
This also demonstrates how parser generators can be expensive code-footprint-wise, as the generated parser class itself is larger than 90KB.
CellularBasic include floating point emulation, which is excluded for the code footprint approximation.

\subsection{JScheme}
JScheme\cite{jscheme} is not directly targeted mobile devices, but it is interesting here, in the sense that it is tiny and elegant implementation of a minimalistic language.
Unfortunately it is also rather slow, TODO:benchmark example.

\subsection{Mojax}
\subsection{Rhino}
Rhino is a JavaScript/EcmaScript implementation on top of the JVM. 
\subsection{Jython}

\section{Other languages}
\subsection{JavaScript/EcmaScript}
\subsection{Python}
\subsection{Forth and other stack languages}
\subsection{Self - prototypes}
\subsection{Scheme/Lisp - minimalistic syntax}
\subsection{Haskell - lazyness}

